Enhanced Material Point Method to face dynamic problems:\thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-13 main.tex
Cambios 12-3-2020

\section{Analysis of the proposed improvements.}
\label{sec:Analysis-proposed-improvements.}

@@

The total simulation time is 0.0002 seconds, a direct consequence of
that is we let the elastic wave generated travel thorough the bar
(from A to C and back to A) at least four times. For the spatial
discretization we will set a mesh size of 3.3325 millimeters. The
initial mesh will be composed by a single particle in the center of
the element. Regarding this, here we give one consideration to
properly modellize this benchmark under the MPM framework. Notice that
the free border of the bar has a horizontal displacement of 0.033
millimeters, therefore a computational domain with an extra gap of
0.033 millimeters is required in order to accommodate the
unconstrained displacement of the particles in the left border of the
bar. Here below the time step is controlled by the courant number, in
this case, we adopt a middle value of it (CFL=0.5) as a compromise
between stability and numerical accuracy.

@@

In order to test the ability of this interpolation technique to deal
with grid crossing instabilities, we will simulate the compression of
a linear elastic soft soil due to the gravity force, \ref{fig:block}.

@@

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/FE_vs_PCE_CFL_05}
  }
  \caption{Geometrical description of a soil block.}
  \label{fig:block-MPM}
\end{figure}

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Block_CFL_01_MPM.png}
  }
  \caption{Geometrical description of a soil block }.}
\label{fig:block-MPM}
\end{figure}

>>>>

\section{Application to linear elasticity dynamic problems.}
\label{sec:Application-linear-elasticity-dynamic-problems}

@@

The total simulation time is 0.0002 seconds, a direct consequence of
that is we let the elastic wave generated travel thorough the bar
(from A to C and back to A) at least four times. For the spatial
discretization we will set a mesh size of 3.3325 millimeters. The
initial mesh will be composed by a single particle in the center of
the element.

Regarding this, here we give one consideration to properly modellize
this benchmark under the MPM framework. Notice that the free border of
the bar has a horizontal displacement of 0.033 millimeters, therefore
a computational domain with an extra gap of 0.033 millimeters is
required in order to accommodate the unconstrained displacement of the
particles in the left border of the bar. Here below the time step is
controlled by the courant number, in this case, we adopt a middle
value of it (CFL=0.5) as a compromise between stability and numerical
accuracy.

@@

In order to test the ability of this interpolation technique to deal
with grid crossing instabilities we will simulate the vertical
compression of a square block (10 by 10 meters) of soft soil sketched
in figure \ref{fig:block} and loaded using an incremental gravitation scheme.

@@

This test was taken from Andersen (2009)\cite{thesis_Andersen_2009}. The
elastic parameters consider for this test are: 
\begin{itemize} 
\item  Initial density : $6\cdot 10^3\ kg/m^3$
\item  Poisson ratio : $0$
\item  Elastic modulus : $5\ MPa$
\end{itemize}
The gravity force is a apply as an external force according to the
equations \eqref{eq:particle_body_forces},
\eqref{eq:nodal_external_forces}. Using a total time period of T (20
seconds) to apply the gravity, it is increased from 0 to 9.81$m/s$
with a sinus function until T/2 and then maintained constant until T
in order to arrive to a state of equilibrium, 
\begin{equation}
  \label{eq:gravity-load-block}
 \mathbf{g}(t) = \left\{
    \begin{array}{ll}
      0.5 \mathbf{g} (\sin(\frac{2t \pi}{T} - \frac{\pi}{2})+1)  & \mbox{if } t \leq T/2 \\
      \mathbf{g} & \mbox{if } t > T/2
    \end{array}
  \right.
\end{equation}

In order to get a stable solution, we will adopt a time step conducted
by a Courant number of 0.1. On the other hand, the explicit predictor-corrector
scheme is here employed looking forward getting better results. For the
initial spatial discretization we will employ four particles per cell
($\Delta x = 2\ m$). The initial setting of particles inside of the
cell changes according to the interpolation technique adopted. For the
bi-linear shape functions (Q4) and the LME shape functions, the initial
position corresponds to the location of the gauss-points in a standard
quadratic finite element. For the uniform-GIMP (uGIMP) shape function
the initial position of each particle is located in the center of each
voxel, due to the fact that in the initial situation, the voxel domain
should not overlap each others, Steffen {\it et al.}
(2008)\cite{Steffen2008}. Figure \ref{fig:vertical-displacement-block}
shows the vertical displacement evolution of a point in the free
surface of the block.
\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Block_CFL_01_Comparative}
  }
  \caption{Comparative of the vertical displacement evolution in a
    point located in the free surface employing different
    interpolation schemes and numerical techniques.} 
  \label{fig:vertical-displacement-block}
\end{figure}

As we can see the simulation using the Q4 interpolation technique
turns out to be unstable and in the second 15 it fails. The uGIMP
simulation is more stable than the one performed by the Q4. Despite
this is still unstable and could trigger severe oscillations if we
pretend to simulate non-linear materials. The LME simulation was
performed using two kinds of shape functions, one with a low value of
the dimensionless parameter, $\gamma = 0.8$, and other with a larger
value of it, $\gamma = 3.0$. Notice that the results are both
stable, but the larger values of $\gamma$ give us a very stable
solution. This is due to the fact that with larger value of $\gamma$,
the shape functions behaves in a similar way to the FEM, which performs
very accurate in those cases with a reasonable mesh distorsion, and
with a lower value it behaves in a similar way to the uGIMP. This
behaviour was noticed previously by \cite{Arroyo2006}, were authors
highlight how by adjusting the spatial variation of $\beta(\vect{x})$,
it is possible to select regions of the domain of analysis which are
treated by finite elements and regions that are treated in the style
of meshfree methods, with seamless transitions between those regions.

Figure \ref{fig:Block-LME3} shows the evolution of the vertical stress
during the loading process. The result is physically realistic as
stress increments linearly from the top to the bottom of the specimen,
and the value of the vertical stress in a material point located in
the bottom of the specimen oscillates centered in $5.2 MPa$, which is
the analytic value given by $\sigma_{yy} = \rho g h_y$.

\begin{figure}
  \centering
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 0 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t0}
    \label{fig:Block-LME3-PCE-t0}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 5 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t025}
    \label{fig:Block-LME3-PCE-t025}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 20 seconds]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t1}
    \label{fig:Block-LME3-PCE-t1}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \caption{Vertical normal stress and position of material points
    during the loading process for a soft soil ($E = 5\ MPa$, $\rho_0
    = 6\cdot 10^3\ kg/m^3$). Numerical parameters considered for the
    simulation are : Local \textit{max-ent} shape function $\gamma =3$
    and explicit PC scheme with CFL 0.1.}
  \label{fig:Block-LME3}
\end{figure}
